Social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{Paper presented at the session 'Health and Mortality in Arctic Populations'. The Ninth International Congress on Arctic Social Sciences, Umeå, June 8--12, 2017}

### 2017-05-26 icass.Rnw
reorganising ...

Figure~\ref{fig:hiscskum} shows the distribution (per cent) of exposure
time over class  for our selected periods. The category of workers is
fairly stable over time. The highest class has
increased substantially, although from a very low level in the 19th
century. The increase of academics in connection to the establishment of
Umeå university explains much of this increase. Within the middle group, there has been a transfer from a majority of farmer during the 19th century to lower white collar from the middle of the 20th century.

<<hiscskum, fig.cap = "Exposure by hisclass (from the top: elite, middle, workers, missing) and time period", fig.height = 4.5>>=
xx <- with(vb0Tab, tapply(exposure, list(period = period, hisclass = hisclass), sum))
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
@

@@

<<nana, results = 'asis'>>=
library(xtable)
##load("data/vb0.rda")
##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
x <- with(vb0, table(period, hisclass, useNA = "ifany"))
x <- round(prop.table(x, 1) * 100)
xtable(x, caption = "Hisclass by time period, per cent.", label = "tab:nana", digits = 0)
@

@@

\section{Results, all causes of death}

@@

% <<hazcalw>>=
% source("R/getPHglm2.R")
% ##nper <- 4
% nper <- length(labb)
% resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
%             nper, labb)
% resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
%             nper, labb)
% @
% 
% 
% <<hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 7>>=
% source("R/plotHaz.glm.R")
% 
% plotHaz.glm(resw[-(1:6)], main = labb, col = scol, layout = c(2, 2))
% @
% 
% <<hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 7>>=
% plotHaz.glm(resm[-(1:6)], main = labb, col = scol, layout = c(2, 2))
% @
% 
% From Figures~\ref{fig:hazplotw} and \ref{fig:hazplotm} it is obvious that
% the idea of proportional hazards is not reasonable. We therefore use another approach.

@@

\subsubsection*{The effect of income and education, 1990--2005}

<<lissyoung>>=
library(eha)
library(xtable)
liss4064 <- loTab[loTab$age < 64.5, ]
@

The statistical significance of the covariates are shown in Table~\ref{tab:liswom}.
<<liswom, results = 'asis'>>=
fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
##cat("Women:\n")
drw <- drop1(fitw1, test = "Chisq")
xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005.", label = "tab:liswom")
##round(summary(fitw1)$coef[-(1:13), ], 4)
@

The estimated effect sizes for women are shown in Figure \ref{fig:liswomfig}.

<<liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005.", fig.height = 4>>=
source("R/plotWEduc.R")
plotWEduc(fitw1)
@

And for men:

<<lismen, results='asis'>>=
fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
##cat("\nMen:\n")
drm <- drop1(fitm1, test = "Chisq")
xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005.")
##round(summary(fitm1)$coef[-(1:13), ], 4)
dcol <- c("blue", "darkgreen", "darkgreen", "red")
@

<<lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005.", fig.height = 4>>=
#source("R/plotWEduc.R")
plotWEduc(fitm1)
@

<<plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013.", fig.height = 4>>=
#oldpar <- par(mfrow = c(1, 2))
#plot(fitw1, fn = "cum", col = dcol, main = "Women")
#plot(fitm1, fn = "cum", col = dcol, main = "Men")
#par(oldpar)
##summary(fitw1)
##summary(fitm1)
@

%See Table~\ref{tab:tablisswom1} the PH regression for women.
<<tablisswom1, results = "asis">>=
#fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
#                 data = liss4064[liss4064$sex == "female", ])
#dr <- drop1(fitwo1, test = "Chisq")
#ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
#    label = "tab:tablisswom1")
@

%See Table~\ref{tab:tablissman1} the PH regression for men.
<<tablissman1, results = "asis">>=
#fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
#                 data = liss4064[liss4064$sex == "male", ])
#dr <- drop1(fitma1, test = "Chisq")
#ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
 #   label = "tab:tablissman1")
@
Both education and income were strongly significant, as well as marital status as expected. Those with the two lowest education level hade the highest mortality, while those with highest level of education had by far the best survival. The differences must be considered large, and it follows almost a continuous improvement according to increasing levels. The same can be said concerning income. Higher incomes result in lower mortality. This is also what could be expected, and conforms to what we know of social health differences in present-day Sweden. Notice however that we cannot definitely establish a causal relationship between income and mortality in particular. Having low incomes may be caused by bad health.

@@

\subsubsection*{The effect of income and education, 1990--2005}

The fact that we find rather unclear associations between HISCLASSES and survival, makes alternative approaches to the question of socio-economic health differences interesting. As for the younger age group, we have included an analysis for the period 1990--2005 with income and education as explanatory variables.

<<lissyoung2>>=
library(eha)
library(xtable)
##load("data/loTab.rda")
##loTab$period <- factor(loTab$period)
liss6589 <- loTab[loTab$age > 64.5, ]
@

The statistical significance of the covariates are shown in
Table~\ref{tab:liswom2}
<<liswom2, results = 'asis'>>=
fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
##cat("Women:\n")
drw2 <- drop1(fitw2, test = "Chisq")
xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005.", label = "tab:liswom2")
##round(summary(fitw1)$coef[-(1:13), ], 4)
@

The estimated effect sizes are shown in Figure \ref{fig:liswomfig2}

<<liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005.", fig.height = 4>>=
#source("R/plotWEduc.R")
plotWEduc(fitw2)
@

And for men:

<<lismen2, results='asis'>>=
fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
##cat("\nMen:\n")
drm2 <- drop1(fitm2, test = "Chisq")
xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005.")
##round(summary(fitm1)$coef[-(1:13), ], 4)
dcol <- c("blue", "darkgreen", "darkgreen", "red")
@

<<lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005.", fig.height = 4>>=
#source("R/plotWEduc.R")
plotWEduc(fitm2)
@

These figures show clear associations both with income and education and for women as well as for men. More years of education resulted in lower mortality. For women the association with additional levels became however smaller at the higher levels, the effect is more visible at the steps above the basic levels. In a similar fashion, increased incomes is associated with improved survival both for men and for women. Apparently, education and income is better to catch up socio-economic differences compared to hisclass, at least for this period.

\section{Results, cardiovascular death}

@@

% \subsubsection*{The effect of income and education, 1990--2005}
% 
% We have also performed the analysis on the alternative approaches on socio-economic differences for the causes of death. It is based on a separate analysis with income and education as explanatory variables.
% 
% <<41lissyoung>>=
% library(eha)
% library(xtable)
% ##load("data/loTab.rda")
% loTab$period <- factor(loTab$period)
% liss4064 <- loTab[loTab$age < 64.5, ]
% @
% 
% 
% The statistical significance of the covariates are shown in
% Table~\ref{tab:41liswom}
% <<41liswom, results = 'asis'>>=
% fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
% ##cat("Women:\n")
% drw <- drop1(fitw1, test = "Chisq")
% xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cardiovascular death.", label = "tab:41liswom")
% ##round(summary(fitw1)$coef[-(1:13), ], 4)
% @
% 
% The estimated effect sizes are shown in Figure \ref{fig:41liswomfig}
% 
% <<41liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
% source("R/plotWEduc.R")
% plotWEduc(fitw1)
% @
% 
% 
% And for men:
% 
% <<41lismen, results='asis'>>=
% fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
% ##cat("\nMen:\n")
% drm <- drop1(fitm1, test = "Chisq")
% xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cardiovascular death.", label = "tab:41lismen")
% ##round(summary(fitm1)$coef[-(1:13), ], 4)
% dcol <- c("blue", "darkgreen", "darkgreen", "red")
% @
% 
% <<41lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitm1)
% @
% 
% 
% <<41plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cardiovascular death.", fig.height = 4>>=
% #oldpar <- par(mfrow = c(1, 2))
% #plot(fitw1, fn = "cum", col = dcol, main = "Women")
% #plot(fitm1, fn = "cum", col = dcol, main = "Men")
% #par(oldpar)
% ##summary(fitw1)
% ##summary(fitm1)
% @
% 
% %See Table~\ref{tab:tablisswom1} the PH regression for women.
% <<41tablisswom1, results = "asis">>=
% #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
% #                 data = liss4064[liss4064$sex == "female", ])
% #dr <- drop1(fitwo1, test = "Chisq")
% #ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
% #    label = "tab:tablisswom1")
% @
% 
% %See Table~\ref{tab:tablissman1} the PH regression for men.
% <<41tablissman1, results = "asis">>=
% #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
% #                 data = liss4064[liss4064$sex == "male", ])
% #dr <- drop1(fitma1, test = "Chisq")
% #ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
%  #   label = "tab:tablissman1")
% @
% 
% The analysis of cardiovascular mortality according to education and income conforms very well to what we find for all-cause mortality, and we will therefore not delve into detail on this issue. More education as well as higher income is advantageous both for women and men for avoiding deaths related to the circulatory system. This is also well in line with our analysis based on HISCLASS above.

@@

% \subsubsection*{The effect of income and education on cardiovascular death, 1990--2005}
% 
% The analysis of cardiovascular mortality for the period 1990--2005 is in this section based on a separate analysis with
% income and education as explanatory variables. The statistical significance of the covariates are shown in Table~\ref{tab:41liswom2} for women and Table~\ref{tab:41lismen2} for men. It confirms that most of the variables are strongly significant except
% 
% <<41lissyoung2>>=
% library(eha)
% library(xtable)
% ##load("data/loTab.rda")
% ##loTab$period <- factor(loTab$period)
% liss6589 <- loTab[loTab$age > 64.5, ]
% @
% 
% 
% 
% <<41liswom2, results = 'asis'>>=
% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
% ##cat("Women:\n")
% drw2 <- drop1(fitw2, test = "Chisq")
% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cardiovascular death.", label = "tab:41liswom2")
% ##round(summary(fitw1)$coef[-(1:13), ], 4)
% @
% 
% 
% <<41lismen2, results='asis'>>=
% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
% ##cat("\nMen:\n")
% drm2 <- drop1(fitm2, test = "Chisq")
% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cardiovascular death.", label = "tab:41lismen2")
% ##round(summary(fitm1)$coef[-(1:13), ], 4)
% dcol <- c("blue", "darkgreen", "darkgreen", "red")
% @
% 
% The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
% 
% <<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitw2)
% @
% 
% 
% And for men:
% 
% 
% 
% <<41lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitm2)
% @
% 
% The pattern is similar to what we find for all-cause mortality. The higher educational level, the lower cardiovascular mortality, and more income leads to better survival. This association is clear and consistent for women as well as for men. A minor difference between men and women is that mortality in the higher levels of education do not lead to lower mortality among women while that is the case among men.

\section{Results, cancer death}

@@

>>>>

Figure~\ref{fig:hiscskum} shows the distribution (per cent) of exposure
time over class  for our selected periods. The category of workers is
fairly stable over time. The highest class has
increased substantially, although from a very low level in the 19th
century. The increase of academics in connection to the establishment of
Umeå University explains much of this increase. Within the middle group, there has been a transfer from a majority of farmer during the 19th century to lower white collar from the middle of the 20th century.

<<hiscskumm, fig.cap = "Exposure by hisclass, men (from the top: elite, middle, workers, missing) and time period", fig.height = 4.5>>=
xx <- with(vb0Tab[vb0Tab$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
@

<<hiscskumw, fig.cap = "Exposure by hisclass, women (from the top: elite, middle, workers, missing) and time period", fig.height = 4.5>>=
xx <- with(vb0Tab[vb0Tab$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
@

@@

<<nanam, results = 'asis'>>=
library(xtable)
##load("data/vb0.rda")
##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
x <- with(vb0[vb0$sex == "male", ], table(period, hisclass, useNA = "ifany"))
x <- round(prop.table(x, 1) * 100)
xtable(x, caption = "Hisclass by time period, men. Per cent.", label = "tab:nana", digits = 0)
@

<<nanaw, results = 'asis'>>=
##library(xtable)
##load("data/vb0.rda")
##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
x <- with(vb0[vb0$sex == "female", ], table(period, hisclass, useNA = "ifany"))
x <- round(prop.table(x, 1) * 100)
xtable(x, caption = "Hisclass by time period, per cent.", label = "tab:nana", digits = 0)
@

@@

\section{The effect of Hisclass on All causes of death}

@@

<<hazcalw>>=
source("R/getPHglm2.R")
##nper <- 4
nper <- length(labb)
resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
            nper, labb)
resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
            nper, labb)
@

% 
% 
% <<hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 7>>=
% source("R/plotHaz.glm.R")
% 
% plotHaz.glm(resw[-(1:6)], main = labb, col = scol, layout = c(2, 2))
% @
% 
% <<hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 7>>=
% plotHaz.glm(resm[-(1:6)], main = labb, col = scol, layout = c(2, 2))
% @
% 
% From Figures~\ref{fig:hazplotw} and \ref{fig:hazplotm} it is obvious that
% the idea of proportional hazards is not reasonable. We therefore use another approach.

@@

% \subsubsection{The effect of income and education, 1986--2013}
% 
% A common alternative to a class analysis is to use income or education for investigating socio-economic health differences. 
% Comparing different approaches are furthermore valuable in order to understand why some are advantaged and others 
% disadvantaged. For the time being, we have to restrict such a comparative analysis to the last decades.
% 
% The analysis for the period 1990--2005 is based on a separate analysis with 
% income and education as explanatory variables and with controls of period 
% and marital status.
% 
% <<lissyoung>>=
% library(eha)
% library(xtable)
% liss4064 <- loTab[loTab$age < 64.5, ]
% con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
% ##load("data/lisaold.rda")
% load(con)
% close(con)
% @
% 
% Both education and income had a strong, statistically significant, effect on survival, as expected. 
% 
% \paragraph{Education, 1990--2013}
% 
% <<educ>>=
% 
% educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
% educ$period <- factor(educ$period)
% educ <- educ[!is.na(educ$educ), ]
% levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
% ##educ$educ <- relevel(educ$educ, ref = "middle")
% educ$civst <- relevel(educ$civst, ref = "married")
% fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
%              data = educ[educ$sex == "male", ])
% fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
%              data = educ[educ$sex == "female", ])
% cols <- c("red", "darkgreen", "black")
% @
% 
% <<ploteduc, fig.cap="Survival by education and sex, 1990--2013.">>=
% oldpar <- par(mfrow = c(1, 2))
% 
% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
% axis(1)
% axis(2, at = c(0, 0.5, 1.0), las = 1)
% abline(h = 0.5, lty = 3)
% box()
% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
% 
% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
% axis(1)
% axis(2, at = c(0, 0.5, 1.0), las = 1)
% abline(h = 0.5, lty = 3)
% box()
% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
% 
% par(oldpar)
% @
% 
% Those with the two lowest education level hade the highest mortality, while those with highest level of education had by far the best survival. The differences must be considered large, and it follows almost a continuous improvement according to increasing levels. 
% 
% \paragraph{Income, 1986--2005}
% 
% <<femeduc>>=
% income <- lisaold[!is.na(lisaold$income), ]
% income$period <- factor(income$period)
% ##income <- income[!is.na(income$income), ]
% ##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
% ##income$income <- relevel(income$income, ref = "low")
% income$civst <- relevel(income$civst, ref = "married")
% fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
%              data = income[income$sex == "male", ])
% fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
%              data = income[income$sex == "female", ])
% cols <- c("darkgreen", "red", "blue", "black")
% 
% @
% 
% <<plotincome, fig.cap="Survival by income and sex, 1986--2005.">>=
% oldpar <- par(mfrow = c(1, 2))
% 
% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
% axis(1)
% axis(2, at = c(0, 0.5, 1.0), las = 1)
% abline(h = 0.5, lty = 3)
% box()
% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
% 
% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
% axis(1)
% axis(2, at = c(0, 0.5, 1.0), las = 1)
% abline(h = 0.5, lty = 3)
% box()
% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
% 
% par(oldpar)
% @
% 
% Higher incomes result in lower mortality. This is also what could be expected, 
% and conforms to what we know of social health differences in present-day Sweden. 
% Notice however that we cannot establish a causal relationship between 
% income and mortality. Bad health, for instance, may be a confounder.
% 
% %%The statistical significance of the covariates are shown in Table~\ref{tab:liswom}.
% %%<<liswom, results = 'asis'>>=
% %%fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
% %%##cat("Women:\n")
% %%drw <- drop1(fitw1, test = "Chisq")
% %%xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005.", label = "tab:liswom")
% %##round(summary(fitw1)$coef[-(1:13), ], 4)
% %@
% 
% %%The estimated effect sizes for women are shown in Figure \ref{fig:liswomfig}.
% 
% %%<<liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005.", fig.height = 4>>=
% %%source("R/plotWEduc.R")
% %%plotWEduc(fitw1)
% %%@
% 
% 
% %%And for men:
% 
% %%<<lismen, results='asis'>>=
% %%fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
% %%##cat("\nMen:\n")
% %%drm <- drop1(fitm1, test = "Chisq")
% %%xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005.")
% %%##round(summary(fitm1)$coef[-(1:13), ], 4)
% %%dcol <- c("blue", "darkgreen", "darkgreen", "red")
% %%@
% 
% %%<<lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005.", fig.height = 4>>=
% %%#source("R/plotWEduc.R")
% %%plotWEduc(fitm1)
% %%@
% 
% % 
% % <<plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013.", fig.height = 4>>=
% % #oldpar <- par(mfrow = c(1, 2))
% % #plot(fitw1, fn = "cum", col = dcol, main = "Women")
% % #plot(fitm1, fn = "cum", col = dcol, main = "Men")
% % #par(oldpar)
% % ##summary(fitw1)
% % ##summary(fitm1)
% % @
% % 
% % %See Table~\ref{tab:tablisswom1} the PH regression for women.
% % <<tablisswom1, results = "asis">>=
% % #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
% % #                 data = liss4064[liss4064$sex == "female", ])
% % #dr <- drop1(fitwo1, test = "Chisq")
% % #ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
% % #    label = "tab:tablisswom1")
% % @
% % 
% % %See Table~\ref{tab:tablissman1} the PH regression for men.
% % <<tablissman1, results = "asis">>=
% % #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
% % #                 data = liss4064[liss4064$sex == "male", ])
% % #dr <- drop1(fitma1, test = "Chisq")
% % #ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
% %  #   label = "tab:tablissman1")
% % @
%

@@

% \subsubsection{The effect of income and education, 1990--2005}
% 
% The fact that we find rather unclear associations between HISCLASSES and survival, makes alternative approaches to the question of socio-economic health differences interesting. As for the younger age group, we have included an analysis for the period 1990--2005 with income and education as explanatory variables.
% 
% <<lissyoung2>>=
% library(eha)
% library(xtable)
% ##load("data/loTab.rda")
% ##loTab$period <- factor(loTab$period)
% liss6589 <- loTab[loTab$age > 64.5, ]
% @
% 
% 
% The statistical significance of the covariates are shown in
% Table~\ref{tab:liswom2}
% <<liswom2, results = 'asis'>>=
% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
% ##cat("Women:\n")
% drw2 <- drop1(fitw2, test = "Chisq")
% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005.", label = "tab:liswom2")
% ##round(summary(fitw1)$coef[-(1:13), ], 4)
% @
% 
% The estimated effect sizes are shown in Figure \ref{fig:liswomfig2}
% 
% <<liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitw2)
% @
% 
% 
% And for men:
% 
% <<lismen2, results='asis'>>=
% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
% ##cat("\nMen:\n")
% drm2 <- drop1(fitm2, test = "Chisq")
% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005.")
% ##round(summary(fitm1)$coef[-(1:13), ], 4)
% dcol <- c("blue", "darkgreen", "darkgreen", "red")
% @
% 
% <<lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitm2)
% @
% 
% These figures show clear associations both with income and education and for women as well as for men. More years of education resulted in lower mortality. For women the association with additional levels became however smaller at the higher levels, the effect is more visible at the steps above the basic levels. In a similar fashion, increased incomes is associated with improved survival both for men and for women. Apparently, education and income is better to catch up socio-economic differences compared to hisclass, at least for this period.

\section{The effect of Hisclass on cardiovascular death}

@@

% \subsubsection{The effect of income and education, 1990--2005}
% 
% We have also performed the analysis on the alternative approaches on socio-economic differences for the causes of death. It is based on a separate analysis with income and education as explanatory variables.
% 
% <<41lissyoung>>=
% library(eha)
% library(xtable)
% ##load("data/loTab.rda")
% loTab$period <- factor(loTab$period)
% liss4064 <- loTab[loTab$age < 64.5, ]
% @
% 
% 
% The statistical significance of the covariates are shown in
% Table~\ref{tab:41liswom}
% <<41liswom, results = 'asis'>>=
% fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
% ##cat("Women:\n")
% drw <- drop1(fitw1, test = "Chisq")
% xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cardiovascular death.", label = "tab:41liswom")
% ##round(summary(fitw1)$coef[-(1:13), ], 4)
% @
% 
% The estimated effect sizes are shown in Figure \ref{fig:41liswomfig}
% 
% <<41liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
% source("R/plotWEduc.R")
% plotWEduc(fitw1)
% @
% 
% 
% And for men:
% 
% <<41lismen, results='asis'>>=
% fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
% ##cat("\nMen:\n")
% drm <- drop1(fitm1, test = "Chisq")
% xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cardiovascular death.", label = "tab:41lismen")
% ##round(summary(fitm1)$coef[-(1:13), ], 4)
% dcol <- c("blue", "darkgreen", "darkgreen", "red")
% @
% 
% <<41lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitm1)
% @
% 
% 
% <<41plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cardiovascular death.", fig.height = 4>>=
% #oldpar <- par(mfrow = c(1, 2))
% #plot(fitw1, fn = "cum", col = dcol, main = "Women")
% #plot(fitm1, fn = "cum", col = dcol, main = "Men")
% #par(oldpar)
% ##summary(fitw1)
% ##summary(fitm1)
% @
% 
% %See Table~\ref{tab:tablisswom1} the PH regression for women.
% <<41tablisswom1, results = "asis">>=
% #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
% #                 data = liss4064[liss4064$sex == "female", ])
% #dr <- drop1(fitwo1, test = "Chisq")
% #ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
% #    label = "tab:tablisswom1")
% @
% 
% %See Table~\ref{tab:tablissman1} the PH regression for men.
% <<41tablissman1, results = "asis">>=
% #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
% #                 data = liss4064[liss4064$sex == "male", ])
% #dr <- drop1(fitma1, test = "Chisq")
% #ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
%  #   label = "tab:tablissman1")
% @
% 
% The analysis of cardiovascular mortality according to education and income conforms very well to what we find for all-cause mortality, and we will therefore not delve into detail on this issue. More education as well as higher income is advantageous both for women and men for avoiding deaths related to the circulatory system. This is also well in line with our analysis based on HISCLASS above.

@@

% \subsubsection{The effect of income and education on cardiovascular death, 1990--2005}
% 
% The analysis of cardiovascular mortality for the period 1990--2005 is in this section based on a separate analysis with
% income and education as explanatory variables. The statistical significance of the covariates are shown in Table~\ref{tab:41liswom2} for women and Table~\ref{tab:41lismen2} for men. It confirms that most of the variables are strongly significant except
% 
% <<41lissyoung2>>=
% library(eha)
% library(xtable)
% ##load("data/loTab.rda")
% ##loTab$period <- factor(loTab$period)
% liss6589 <- loTab[loTab$age > 64.5, ]
% @
% 
% 
% 
% <<41liswom2, results = 'asis'>>=
% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
% ##cat("Women:\n")
% drw2 <- drop1(fitw2, test = "Chisq")
% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cardiovascular death.", label = "tab:41liswom2")
% ##round(summary(fitw1)$coef[-(1:13), ], 4)
% @
% 
% 
% <<41lismen2, results='asis'>>=
% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
% ##cat("\nMen:\n")
% drm2 <- drop1(fitm2, test = "Chisq")
% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cardiovascular death.", label = "tab:41lismen2")
% ##round(summary(fitm1)$coef[-(1:13), ], 4)
% dcol <- c("blue", "darkgreen", "darkgreen", "red")
% @
% 
% The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
% 
% <<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitw2)
% @
% 
% 
% And for men:
% 
% 
% 
% <<41lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitm2)
% @
% 
% The pattern is similar to what we find for all-cause mortality. The higher educational level, the lower cardiovascular mortality, and more income leads to better survival. This association is clear and consistent for women as well as for men. A minor difference between men and women is that mortality in the higher levels of education do not lead to lower mortality among women while that is the case among men.

\section{The effect of Hisclass on cancer death}

@@

\section{The effect of Education on All causes of death, 1990--2013}

<<lissyoung3>>=
library(eha)
library(xtable)
liss4064 <- loTab[loTab$age < 64.5, ]
con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
##load("data/lisaold.rda")
load(con)
close(con)
@

Both education and income had a strong, statistically significant, effect on survival, as expected.

<<educ>>=
